---
title: "Exploring Spotify Music Data"
author: "By Mwangi George"
date: "Last edited `r format(Sys.Date(), '%b, %d %Y')`"
output: 
    github_document:
      toc: yes
---


```{r include=FALSE, echo=FALSE}
# loading packages
pacman::p_load(tidyverse, data.table, janitor, ggthemes)

# laoding dataset
spotify <- read_csv("datasets/top10s.csv", show_col_types = F) %>% clean_names()

metadata <- read_csv("datasets/spotify_metadata.csv", show_col_types = F)

spotify %>% 
  group_by(year) %>% 
  summarise(n())
names(metadata)
names(spotify)
metadata

# rename variable names
spotify %>%
  select(
    id = x1,
    song_title = title,
    song_artist = artist,
    top_genre,
    year,
    beats_per_min = bpm,
    song_energy = nrgy,
    danceability = dnce,
    song_loudness = d_b,
    live,
    valence = val,
    duration = dur,
    acousticness =acous,
    speechiness = spch,
    popularity = pop
  ) -> spotify

# remove duplicates and check for NA's
spotify %>% 
  distinct() %>% 
  filter(complete.cases(.)) -> spotify

# distinct top genres

spotify %>% 
  distinct(top_genre)

spotify %>% 
  glimpse()

spotify %>% 
  group_by(year) %>% 
  summarise(mean_speechiness = mean(speechiness)) %>% view()

spotify %>% 
  count(year, speechiness) %>% 
  arrange(desc(speechiness))
summary(spotify$speechiness)

spotify %>% 
  group_by(year) %>% 
  summarise()
```

# Amount of Spoken Words

This section aims to explore the amount of spoken words in the spotify dataset. This variable is represented by the speechiness column, where more values mean more spoken words. Lets start by understanding the general statistics about the variable by calculating the overall mean, median and other statistics as follows. 
```{r}
summary(spotify$speechiness)
```

From the above output, we can see that, all the songs from 2010 to 2019 have an average speechiness score of 8.36 and a median score of 5. The minimum and maximum speechiness scores are 0 and 48 respectively. There is quite a big difference between the mean and the median, which means there are outliers in the variable. Let's utilize a density plot to explore visualize the distribution of the overall speechiness of the songs.
```{r}
spotify %>% 
  ggplot(aes(speechiness, group =1))+
  geom_density(fill = "blue", alpha = .4)+
  theme_clean() +
  labs(
    title = "Overall Mean Speechiness",
  )
```


The distribution of speechiness is heavily skewed to the right. This explains why the mean is greater than than the median. This skewness renders the mean a poor measure of centrality. We should therefore use the median as a safer alternative to describe centrality of the speechiness variable.
Let's therefore explore how the median speechiness of the songs varied over time. 
```{r}
spotify %>%
  group_by(year) %>%
  summarise(median_speechiness = median(speechiness)) %>%
  ggplot(aes(factor(year), median_speechiness, group = 1)) +
  geom_line() +
  geom_point() +
  expand_limits(y = c(4, 10)) +
  theme_clean() +
  labs(
    title = "Median Speechiness over time",
    x = "Year",
    y = "Median Speechiness"
  )
```

According to the above graph, from 2010 to 2015, the median level of speechiness remained at 5. In 2016, this value increased to 6, and it stayed that way until 2018, when it dropped back to 5.

Let's explore the association between speechiness and popularity of the songs. 
```{r}
# correlation coefficient
cor(spotify$speechiness, spotify$popularity)
```
The correlation coefficient between the variables speechiness and popularity is -0.0415. This value means that there exists a weak negative association between speechiness and popularity. More words in a song mean is associated with reduced popularity but this association is very weak. We can visualize this relationship using a scatter plot. 

```{r}
spotify %>%
  ggplot(aes(speechiness, popularity)) +
  geom_jitter(color = "blue", alpha = .4) +
  theme_clean() +
  labs(
    title = "Popularity Vs Speechiness",
    x = "Speechiness",
    y = "Popularity"
  )
```


# Loudness
In this part, we are going to explore Loudness of the songs. Let's start by calculating summary statistics of variable.
```{r}
summary(spotify$song_loudness)
```
On average, a song has an average loudness of -5.579 decibel. The minimum and maximum loudness values are -60 decibel and -2 decibel respectively. We can utilize a histogram to visualize the distribution.
```{r}
spotify %>% 
  ggplot(aes(song_loudness))+
  geom_histogram(bins = 30)+
  theme_clean()+
  labs(title = "Distribution of song loudness")
```
```{r}
spotify %>% 
  filter(song_loudness == -60)
```

We learn that most of the values lie between -12 dB and -5 dB. There is one song, however, "Million Years Ago" by Adele,  with -60 dB. Notice that this is the same identified with zero words. We can take this analysis further by exploring the correlation between loudness and popularity of a song.
```{r}
# correlation coeffient between song loudness and popularity
cor(spotify$song_loudness, spotify$popularity)
```
 The output suggests shows that there is a weak positive association between the loudness of a song and its popularity. More loudness in a song is associated with more popularity. Let's visualize this relationship with a scatter plot.
```{r}
spotify %>% 
  mutate(year = as.factor(year)) %>% 
  ggplot(aes(song_loudness, popularity))+
  geom_jitter(alpha = .5)+
  facet_wrap(~ year)+
  labs(title = "Popularity Versus Loudness")
```

The relationship is not very clear but we can see that in 2019, most songs had higher sound and were associated with high popularity compared to 2016 where some songs had low sound and were associated with low popularity.




































